Formalization of Network Flow Algorithms

### 2017-03-02 paper/root.tex
remark about overflows, fixed overful hbox (listing rightmargin setting caused spurious overful hbox warnings)

\input{lstisabelle}
\lstset{basicstyle=\footnotesize\ttfamily\slshape}
\lstset{captionpos=b}
\lstset{numberbychapter=false}
\lstset{autogobble=true}
\lstset{xleftmargin=.15in, xrightmargin=.15in}

%%%%%%%%% IMPORTANT %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%change both above and following margins to have consistent indentation
\newenvironment{lstquote}%
  {\list{}{\leftmargin=0.15in\rightmargin=0.15in}\item[]}%
  {\endlist}

@@

% \overfullrule=8pt

@@

\subsection{Formalization of the Generic Algorithm}
For the complexity reasoning, we will count the different types of operations. Thus, it is convenient to formalize the algorithm as a labeled transition system
over flows and heights, the labels indicating the performed operation:
\begin{lstlisting}
  inductive_set pr_lts :: "((preflow\<times>labeling) \<times> op_ty \<times> (preflow\<times>labeling)) set" 
  where
    "push_pre f l e ==> ((f,l),PUSH,(push_eff f e,l)) \<in> pr_lts"
  | "relabel_pre f l u ==> ((f,l),RELABEL,(f,relabel_eff f l u)) \<in> pr_lts"
\end{lstlisting}  
Here, the \isai{inductive_set} construction defines the least relation that can be derived by the specified rules. 
A push operation can be implemented in time $O(1)$, (only the flow on a single edge has to be modified) and a relabel operation can be implemented in time $O(V)$.
(to find the new height, we have to consider the heights of all successor nodes in the residual graph) We define 
\isai{cost PUSH == 1 | cost RELABEL == card(V)} and prove the following theorem:
\begin{lstlisting}
theorem (in Network) generic_preflow_push_OV2E_and_correct:
  assumes "((f_0, l_0), p, (f, l)) \<in> pr_lts^*" 
  shows "(\<sum>x<-p. cost x) \<le> 26 * (card V)$^2$ * card E"
    and "(f,l)\<notin>Domain pr_lts --> isMaxFlow f"
\end{lstlisting}  
where \isai$(f_0,l_0)$ is the initial flow and height labeling, and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}. 
Intuitively, the cost of any (partial) run \isai$p$ from the initial state to \isai$(f,l)$ is bounded by $O(V^2E)$, and if the 
run is complete, \ie no more operations apply (\isai$(f,l)\<notin>Domain pr_lts$), then the resulting flow is maximal.

@@

  We implement capacities by (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers. Thus, we could chose any executable number representation here.}. Note that an implementation by fixed precision numbers would also be possible,
  but requires additional checks on the network to ensure that no overflows can happen. 
  % TODO: Can we trick the Java reference implementation on integer overflow?
  Nodes are represented by natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
  are realized as arrays of size $N\times N$ with row-major indexing, such that the successors of a node are close together in memory.
  The adjacency map of the network is implemented by an array of lists of nodes. 
  An augmenting path is represented by a list of edges, \ie a list of pairs of nodes. 
  The height frequency map and the maps from nodes to height labels and excess values are implemented by arrays.
  The queue of the FIFO push-relabel algorithm and the list of the relabel-to-front algorithm are implemented by two stacks, allowing for amortized constant time operations.

>>>>

\input{lstisabelle}
\lstset{basicstyle=\footnotesize\ttfamily\slshape}
\lstset{captionpos=b}
\lstset{numberbychapter=false}
\lstset{autogobble=true}
\lstset{xleftmargin=.15in, xrightmargin=.05in}

%%%%%%%%% IMPORTANT %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%change both above and following margins to have consistent indentation
\newenvironment{lstquote}%
  {\list{}{\leftmargin=0.15in\rightmargin=.05in}\item[]}%
  {\endlist}

@@

\overfullrule=8pt

@@

\subsection{Formalization of the Generic Algorithm}
For the complexity reasoning, we will count the different types of operations. Thus, it is convenient to formalize the algorithm as a labeled transition system
over flows and heights, the labels indicating the performed operation:
\begin{lstlisting}
  inductive_set pr_lts :: "((preflow\<times>labeling) \<times> op_ty \<times> (preflow\<times>labeling)) set"
  where
    "push_pre f l e ==> ((f,l),PUSH,(push_eff f e,l)) \<in> pr_lts"
  | "relabel_pre f l u ==> ((f,l),RELABEL,(f,relabel_eff f l u)) \<in> pr_lts"
\end{lstlisting}  
Here, the \isai{inductive_set} construction defines the least relation that can be derived by the specified rules. 
A push operation can be implemented in time $O(1)$, (only the flow on a single edge has to be modified) and a relabel operation can be implemented in time $O(V)$.
(the new height depends on the heights of the successors in the residual graph) We define 
\isai{cost PUSH == 1 | cost RELABEL == card(V)} and prove the following theorem:
\begin{lstlisting}
theorem (in Network) generic_preflow_push_OV2E_and_correct:
  assumes "((f_0, l_0), p, (f, l)) \<in> pr_lts^*" 
  shows "(\<sum>x<-p. cost x) \<le> 26 * (card V)$^2$ * card E"
    and "(f,l)\<notin>Domain pr_lts --> isMaxFlow f"
\end{lstlisting}  
where \isai$(f_0,l_0)$ is the initial flow and height labeling, and \isai{pr_lts^*} denotes the reflexive transitive closure of \isai{pr_lts}. 
Intuitively, the cost of any (partial) run \isai$p$ from the initial state to \isai$(f,l)$ is bounded by $O(V^2E)$, and if the 
run is complete, \ie no more operations apply (\isai$(f,l)\<notin>Domain pr_lts$), then the resulting flow is maximal.

@@

  We implement capacities by (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers. Thus, we could chose any executable number representation here.}. Note that an implementation by fixed precision numbers would also be possible,
  but requires additional checks on the network to ensure that no overflows can happen.\footnote{Actually, the unverified reference 
  implementations of the push-relabel algorithms~\cite{ChGo97,ICPC-notebook} that we use in our benchmarks (\cf Section~\ref{sec:benchmarks}) lack such checks, and silently report erroneous maximum flow values in case of overflow.}
  % TODO: Can we trick the Java reference implementation on integer overflow?
  Nodes are represented by natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
  are realized as arrays of size $N\times N$ with row-major indexing, such that the successors of a node are close together in memory.
  The adjacency map of the network is implemented by an array of lists of nodes. 
  An augmenting path is represented by a list of edges, \ie a list of pairs of nodes. 
  The height frequency map and the maps from nodes to height labels and excess values are implemented by arrays.
  The queue of the FIFO push-relabel algorithm and the list of the relabel-to-front algorithm are implemented by two stacks, allowing for amortized constant time operations.
